---
title: "COVID19 Cases in Real time"
author: "Alejandro Granados"
date: "23/03/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The dataset: Time series data for COVID19 worldwide. Updated daily. 

<br><br>

This analysis builds up on the amazing work and ideas by  [Tomas Pueyo](https://medium.com/@tomaspueyo/coronavirus-act-today-or-people-will-die-f4d3d9cd99ca). Pueyo's used a [public dataset](https://github.com/CSSEGISandData/COVID-19/blob/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Confirmed.csv) compiled by John Hopkins University. 



```{r loadlib,echo=F, results='hide', message=F, warning=F}

library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)
library(RCurl)
library(ggthemes)
library(ggrepel)
library("rnaturalearth")
library("rnaturalearthdata")
library("sf")
library(rgeos)
library(tmap)
library(RColorBrewer)
library(cowplot)
library(plotly)

```


```{r data_url, echo=F, results='hide', message=F, warning=F}
covid_url = getURL("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv") 
#recovered_url = getURL("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Recovered.csv")
deaths_url = getURL("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv")



# When reading the raw data files, the dates come as columns and are not properly formatted
# This function will convert a string from the raw format into the format dd-mm-yyyy which can be read and converted to a Date object
convertToDate2<-function(x){     

  month_day=do.call(rbind,str_split(x,"\\."))
  month_day[,1] %>% str_replace(pattern ="X",replacement="") ->month
  month_day[,3] %>% str_replace(pattern ="20",replacement="2020") ->year

  return(paste(year, month, month_day[,2],sep="-"))
}

# Load files
covid_data = read.csv(text = covid_url)
covid_deaths = read.csv(text = deaths_url)
#covid_recovered = read.csv(text = recovered_url)

# Convert the date format
names(covid_data)[-c(1:4)] = convertToDate2(names(covid_data)[-c(1:4)])
names(covid_deaths)[-c(1:4)] = convertToDate2(names(covid_deaths)[-c(1:4)])
#names(covid_recovered)[-c(1:4)] = convertToDate2(names(covid_recovered)[-c(1:4)])

# Use gather to make tidy data frames
#covid_recovered %>% gather('Date','Cases',-c(Province.State,Country.Region,Lat,Long)) ->tidy_recovered
covid_data %>% gather('Date','Cases',-c(Province.State,Country.Region,Lat,Long)) ->tidy_cases
covid_deaths %>% gather('Date','Cases',-c(Province.State,Country.Region,Lat,Long)) ->tidy_deaths

# Convert date to Date objects (easier to use)
tidy_cases$Date = as.Date(tidy_cases$Date)
tidy_deaths$Date = as.Date(tidy_deaths$Date)
#tidy_recovered$Date = as.Date(tidy_recovered$Date)

# Rename our variables of interest
names(tidy_deaths)[6]<-'Deaths'
#names(tidy_recovered)[6]<-'Recovered'

# Merge into single dataset 
#covid19_dataset <- merge(merge(tidy_cases,tidy_deaths), tidy_recovered) 
covid19_dataset <- merge(tidy_cases,tidy_deaths)


```

## Fig1. World distribution of COVID19 cases 
To visualize the dataset as a world heatmap we first summarise each country as their current number of confirmed cases. We can use the library rnaturalearth to retrieve a dataframe with geographic information about all countries plus additional information such as estimated population, GDP, etc. 

```{r echo=F, results='hide', message=F, warning=F,fig.height = 8, fig.width = 8}


# Retrive World information 
world <- ne_countries(scale = "medium", returnclass = "sf")
#world_moll = st_transform(world, crs = "+proj=moll")
world_moll = lwgeom::st_transform_proj(world,crs='+proj=wintri')
# get the total (max) cases per country for map heatmap
# For data with Recovered
# covid19_dataset %>%  group_by(Country.Region,Date) %>% 
#                     summarize(Cases = sum(Cases),Deaths=sum(Deaths),Recovered = sum(Recovered))  %>% 
#                     group_by(Country.Region) %>% 
#                     summarize(Cases = max(Cases),Deaths = max(Deaths),Recovered = max(Recovered)) %>% 
#                     arrange(desc(Cases)) -> static_data
#with out Recovered
covid19_dataset %>%  group_by(Country.Region,Date) %>% 
                    summarize(Cases = sum(Cases),Deaths=sum(Deaths))  %>% 
                    group_by(Country.Region) %>% 
                    summarize(Cases = max(Cases),Deaths = max(Deaths)) %>% 
                    arrange(desc(Cases)) -> static_data

# Some names for countries don't agree between the COVID data and the World data so we replace them
static_data$Country.Region %>% str_replace("US","United States") %>% 
                            str_replace("Korea, South","Korea") -> static_data$Country.Region

# 'name' will be the variable for the country's name
names(static_data)[1] = 'name'
# merge dataset by 'name'
left_join(world_moll, static_data, by='name') -> world_covid


world_covid %>% mutate(COVID_cases_log = log10(Cases)) %>% 
                tm_shape() + 
                tm_fill(col="Cases",breaks = 10^(0:5),
                        title = paste("COVID19 cases",format(Sys.Date(),'%b/%d')), 
                        legend.format =list(format="f", big.mark = ',',scientific = T),style ="cont") + 
                tm_borders()  + 
                tm_layout(legend.outside = F,legend.text.size = 0.7,legend.title.size = 0.7)


```
<br><br>
<br><br>

```{r align_fun, echo=F, results='hide', message=F, warning=F}
# # Group data by country and ADD the numbers (we loose the information by State, City)
# covid19_dataset %>% group_by(Country.Region,Date) %>% 
#                     summarise( Cases = sum(Cases), Deaths = sum(Deaths),Recovered = sum(Recovered) ) -> covid19_countries
# without Recovered
covid19_dataset %>% group_by(Country.Region,Date) %>% 
                     summarise( Cases = sum(Cases), Deaths = sum(Deaths)) -> covid19_countries


# rename the variable for the country's name (for compatibility with the World dataset)
names(covid19_countries)[1] = 'name'
# Fix the names (for compatibility with the World dataset)
covid19_countries$name%>% str_replace("US","United States") %>% str_replace("Korea, South","Korea") -> covid19_countries$name

# Join the COVID19 data with World data for normalization by country's population 
left_join(covid19_countries, world_moll %>% select(name,pop_est,region_wb),by='name') -> covid19_countries

# Normalize Cases and Deaths by million inhabitants
covid19_countries %>% mutate(Cases_pm = Cases / (pop_est/10^6) ) -> covid19_countries
covid19_countries %>% mutate(Death_pm = Deaths / (pop_est/10^6) ) -> covid19_countries


alignData<-function(covid19_countries, by='Cases', since=100, plot_countries  = c()){
  covid19_countries %>% group_by(name) %>% filter(name %in% plot_countries) -> x
    if(by=='Cases_pm'){
        x %>% filter(Cases_pm >=since) %>% top_n(1,-Cases_pm) %>% group_by(name) %>% top_n(1,Date) -> Dates_of_1Casepm
    }else if(by=='Deaths'){
        x %>% filter(Deaths >=since) %>% top_n(1,-Deaths) %>% group_by(name) %>% top_n(1,Date) -> Dates_of_1Casepm
    }else if(by =="Cases"){
      x %>% filter(Cases >=since) %>% top_n(1,-Cases) %>% group_by(name) %>% top_n(1,Date) -> Dates_of_1Casepm
    }

    aligned_data = covid19_countries

    aligned_data[,"Diff"] <- NA
    for(i in 1:dim( Dates_of_1Casepm )[1] ){
       aligned_data[aligned_data$name==Dates_of_1Casepm$name[i],]$Diff =  aligned_data$Date[aligned_data$name==Dates_of_1Casepm$name[i]] - Dates_of_1Casepm$Date[i]   
    }

    if(by=='Cases_pm'){
        return(aligned_data %>%  filter(name %in% plot_countries) %>% filter(Cases_pm >=since))
    }else if(by=='Deaths'){
        return(aligned_data %>%  filter(name %in% plot_countries) %>% filter(Deaths >since))
    }else if(by=='Cases'){
        return(aligned_data %>%  filter(name %in% plot_countries) %>% filter(Cases >since))
    }
}


```


```{r theme_def, echo=F, results='hide', message=F, warning=F}

#let's build a theme
 my_theme<-theme_bw() + theme(text = element_text(size=20)) + 
    theme(panel.grid.minor = element_blank(), 
        panel.grid.major = element_line(color = "gray50", size = 0.5),
        panel.grid.major.x = element_blank(),
        panel.background = element_blank(),
        line = element_blank(),
        panel.border = element_blank()) + 
        theme(text = element_text(size = 20),
        legend.title = element_blank(), 
        legend.position = 'top',
        plot.title = element_text(hjust = -0.15,
                                  vjust = 2.12,
                                  colour = "black",
                                  size =14,face='bold')) + theme(legend.text = element_text(size=12))

```



## Fig 2. Cases per million habitants for countries with more than 1k cases 
After filtering for countries with at least 1000 cases, we can plot the data relative to the day when different countries passed the 1 COVID-19 case per million habitants. 


```{r prep_data ,echo=F, results='hide', message=F, warning=F}
# Filter for countries with at least 1000 cases 

static_data %>% filter(Cases >=1000) %>% pull(name) -> plot_countries
static_data %>% filter(Deaths >=20) %>% pull(name) -> plot_countries_deaths


# Align relative to the day of the 1 case p/million 
aligned_test = alignData(covid19_countries, by="Cases_pm",since = 1,plot_countries = plot_countries)
aligned_deaths = alignData(covid19_countries, by ="Deaths",since = 10, plot_countries = plot_countries_deaths)
```

```{r Fig2, echo=F, results='hide', message=F, warning=F,fig.height = 8, fig.width = 8 }
colorCount = length(unique(world_moll$region_wb))
getPalette = colorRampPalette(brewer.pal(8, "Accent"))
colors = getPalette(colorCount)
colors = colors[c(2,1,3,6,5,4,7,8)]


# for color purposes 
aligned_test$region_wb <- factor(aligned_test$region_wb, levels = c('Europe & Central Asia','East Asia & Pacific','Latin America & Caribbean','North America','Middle East & North Africa') )

aligned_test %>% ggplot(aes(x =Diff,y = Cases_pm,color=region_wb,group=name)) + 
    geom_path(alpha = 0.6) + geom_point(size = 1)  + 
    geom_text_repel(data =aligned_test %>% filter(Date == last(Date)), 
                    aes(label = unique(aligned_test$name)), 
                    nudge_x =2, 
                    xlim = c(max(aligned_test$Diff[aligned_test$name=='Korea']) + 7,NA) , 
                    hjust = 0, direction ='y',segment.size = 0.1, 
                    segment.colour = 'gray50',size = 4)  + 
    xlim(0,max(aligned_test$Diff)+4)  + 
    scale_y_continuous(trans = 'log10',name = "Cases per million habitants") + 
    scale_color_manual(values= colors) + 
    guides(col = guide_legend(nrow = 2)) + 
    xlab("Days since 1 case p million") + ggtitle("COVID19 Cases per million habitants\n") + 
  my_theme-> p_confirmed  
    p_confirmed_1 = add_sub(p_confirmed,"Data: 2019 Novel Coronavirus Visual Dashboard; John Hopkins University CSSE",
                x = -0.07,
                hjust = 0,
                fontface = "plain",
                size= 10) 
    ggdraw(p_confirmed_1)

```



<br><br><br><br>

## Fig 3. Deaths since day of 10th death 


```{r Fig3, echo=F, results='hide', message=F, warning=F,fig.height = 8, fig.width = 8 }


# for color purposes 
aligned_deaths$region_wb <- factor(aligned_deaths$region_wb, levels = c('Europe & Central Asia','East Asia & Pacific','Latin America & Caribbean','North America','Middle East & North Africa') )


aligned_deaths %>% ggplot(aes(x =Diff,y = Deaths,color=region_wb,group=name)) + 
    geom_path(alpha = 0.6) + geom_point(size = 1)  + 
    geom_text_repel(data =aligned_deaths %>% filter(Date == last(Date)), 
                    aes(label = unique(aligned_deaths$name)), 
                    nudge_x =2, 
                    xlim = c(max(aligned_deaths$Diff[aligned_deaths$name=='Korea']) + 7,NA) , 
                    hjust = 0, direction ='y',segment.size = 0.1, 
                    segment.colour = 'gray50',size = 4)  + 
    xlim(0,max(aligned_deaths$Diff)+4)  + 
    scale_y_continuous(trans = 'log10',name = "Number of deaths") + 
    annotation_logticks(sides = 'l') +
    scale_color_manual(values= colors) + 
    guides(col = guide_legend(nrow = 2)) + 
    xlab("Days since 10th death") + ggtitle("Deaths by COVID19\n") + 
  my_theme-> p_deaths  
    p_deaths_1 = add_sub(p_deaths,"Data: 2019 Novel Coronavirus Visual Dashboard; John Hopkins University CSSE",
                x = -0.07,
                hjust = 0,
                fontface = "plain",
                size= 10) 
    ggdraw(p_deaths_1)

```


<br><br><br><br>

### Interactive versions

<br><br>

```{r Fig2A, echo=F, message=F, warning=F,fig.height = 6, fig.width = 8 }

ggplotly(p_confirmed + xlim(0,40))

```

<br><br>
```{r Fig3A, echo=F, message=F, warning=F,fig.height = 6, fig.width = 8 }

ggplotly(p_deaths + xlim(0,40))

```



<br><br>

## Fig 4. Death vs Confirmed cases per million habitants
<br><br>
```{r Fig4, echo=F, results='hide', message=F, warning=F,fig.height = 8, fig.width = 8 }

covid19_countries %>% filter(Deaths >2, Cases>300) %>% filter(!is.na(region_wb))->aligned_scatter
aligned_scatter$region_wb <- factor(aligned_scatter$region_wb, levels = c('Europe & Central Asia','East Asia & Pacific','Latin America & Caribbean','North America','Middle East & North Africa','South Asia') )

aligned_scatter %>% filter(Date ==last(Date)) %>% ggplot(aes(x =Cases_pm, y = Death_pm)) +  
    geom_point(aes(color = region_wb),shape = 21, fill ='white',size = 3,stroke = 1.5) + 
    scale_color_manual(values = colors) + 
    scale_y_continuous(trans = 'log10',name = "Deaths per million habitants",labels=scales::comma) + 
    scale_x_continuous(trans = 'log10',name = "Confirmed cases per million habitants",labels=scales::comma) + 
    annotation_logticks(sides = 'lb') +
    geom_smooth(method ='lm',formula = y ~x,se  =F,color='salmon3') +
    geom_text_repel(data = aligned_scatter %>% filter(Date==last(Date)),
                    aes(label = unique(aligned_scatter$name)), 
                    segment.size =0.1) + 
    ggtitle(paste("COVID19 confirmed cases and deaths, ",format(Sys.Date(),'%b/%d'),sep="") )+ 
    my_theme
  
    
        
```
    
    
<br><br>

## Fig 5. Deaths per million habitants since the 100th case was detected

<br><br>


```{r Fig5, echo=F, results='hide', message=F, warning=F,fig.height = 8, fig.width = 8}

static_data %>% filter(Cases >=1000) %>% pull(name) -> plot_countries_1000case


# Align relative to the day of the 1 case p/million 
aligned_300_case = alignData(covid19_countries, by="Cases",since = 100,plot_countries = plot_countries_1000case)

aligned_300_case %>% filter(!is.na(region_wb)) -> aligned_300_case
aligned_300_case$region_wb <- factor(aligned_300_case$region_wb, levels = c('Europe & Central Asia','East Asia & Pacific','Latin America & Caribbean','North America','Middle East & North Africa','South Asia') )



aligned_300_case  %>% filter(Date ==last(Date)) %>% ggplot(aes(x=Diff,y = Death_pm)) + 
 geom_point(aes(color = region_wb),shape = 21, fill ='white',size = 3,stroke = 1.5) + 
    scale_color_manual(values = colors) + 
    scale_y_continuous(trans = 'log10',name = "Deaths per million habitants",labels=scales::comma) + 
    scale_x_continuous(trans = 'log10',name = "Days since the 100th confirmed case",labels=scales::comma) + 
    annotation_logticks(sides = 'lb') +
  guides(col = guide_legend(nrow = 2)) + 
    geom_text_repel(data = aligned_300_case %>% filter(Date==last(Date)),
                    aes(label = unique(aligned_300_case$name)), 
                    segment.size =0.1) + 
    ggtitle(paste("COVID19 deaths since day of 100th confirmed case, ",format(Sys.Date(),'%b/%d'),sep="") )+ my_theme


```