---
title: "COVID19 Cases in Real time"

knit: (function(input_file, encoding) {
  out_dir <- '.';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'COVID19.html'))})

author: "Alejandro AG"
date: "18/03/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r loadlib, echo=F, results='hide', message=F, warning=F}
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)
library(RCurl)
library(ggthemes)
library(ggrepel)
library("rnaturalearth")
library("rnaturalearthdata")
library("sf")
library(rgeos)
library(tmap)
```

## The dataset: Time series data for COVID19 cases in xx countries. Updated daily. 

<br><br>

This analysis build up on the amazing work and ideas by  [Tomas Pueyo](https://medium.com/@tomaspueyo/coronavirus-act-today-or-people-will-die-f4d3d9cd99ca). Pueyo's used a [public dataset](https://github.com/CSSEGISandData/COVID-19/blob/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Confirmed.csv) compiled by John Hopkins University. 



**Downloading the datasets:**


```{r echo=F, results='hide', message=F, warning=F}
aa = getURL("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Confirmed.csv")

covid_data = read.csv(text = aa)

covid_data %>% group_by(Country.Region) %>% select(-c(Lat,Long,Province.State)) -> filtered_data



# # #
deaths_url = getURL("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Deaths.csv")

covid_deaths = read.csv(text = deaths_url)
covid_deaths %>% group_by(Country.Region) %>% select(-c(Lat,Long,Province.State)) -> filtered_deaths


```


```{r echo=F, message = F, warning = F}
# Format the dataset, specially the date 
convertToDate2<-function(x){     

  month_day=do.call(rbind,str_split(x,"\\."))
  month_day[,1] %>% str_replace(pattern ="X",replacement="") ->month
  month_day[,3] %>% str_replace(pattern ="20",replacement="2020") ->year

  return(paste(year, month, month_day[,2],sep="-"))
}

x = names(filtered_data)[-1];
names(filtered_data) = c(names(filtered_data)[1], convertToDate2(x) )

filtered_data %>% gather('Date','Cases',-c(Country.Region)) -> tidy_data

#now data is tidy 
x = names(filtered_deaths)[-1];
names(filtered_deaths) = c(names(filtered_deaths)[1], convertToDate2(x) )

filtered_deaths %>% gather('Date','Cases',-c(Country.Region)) -> tidy_deaths



```
<br><br>


## Total cases COVID19 worldwide


```{r echo=F, message = F, warning = F, fig3, fig.height = 6, fig.width = 12 }
# same countries as before
# we need to rename manually some of them because they don't appear with the same name as in the data frame

theme_set(theme_bw())

world <- ne_countries(scale = "medium", returnclass = "sf")

 tidy_data %>%  group_by(Country.Region,Date) %>% summarize(Cases = sum(Cases))  %>% group_by(Country.Region) %>% summarize(max = max(Cases)) %>% arrange(desc(max)) %>% select(Country.Region,max) -> top_countries


top_countries$Country.Region %>% str_replace("US","United States") %>% str_replace("Korea, South","Korea") -> long_names

top_countries$Country.Region = long_names
# we are going to do left-join by "name"
# we can also add more variables at this point, time to 100th case, death rate, etc
names(top_countries)<-c("name","Covid_cases")

world_moll = st_transform(world, crs = "+proj=moll")

left_join(world_moll,top_countries, by ="name") -> world_covid


world_covid %>% mutate(COVID_cases_log = log10(Covid_cases)) %>% tm_shape() + tm_fill(col="Covid_cases",breaks = 10^(0:5),title = paste("COVID19 cases",format(Sys.Date(),'%b/%d')), legend.format =list(format="f", big.mark = ',',scientific = T),style ="cont") + tm_borders()  + tm_layout(legend.outside = F,legend.text.size = 1.5,legend.title.size = 2)
```

<br><br>


## Total cases per day (countries with more than 1000 cases as today)

```{r echo=F, message = F, warning = F}
tidy_data %>% group_by(Country.Region,Date) %>% summarize(Cases = sum(Cases)) %>% filter(Cases ==max(Cases)) %>% arrange(desc(Cases)) %>% filter(Cases>=1000) %>% select(Country.Region) -> plot_countries

tidy_data$Date <-as.Date(tidy_data$Date)

tidy_data %>% group_by(Country.Region,Date) %>% summarize(Cases = sum(Cases)) %>% filter(Country.Region %in% plot_countries$Country.Region) %>% filter(Country.Region!="China") %>% filter(Date>as.Date('2020-02-14')) -> plot_data 

plot_data %>% filter(Date==last(Date)) %>% pull(Country.Region) %>% as.character() -> label_countries 

label_countries %>% str_replace(pattern = "United Kingdom","UK") %>% str_replace("Korea, South","S Korea") -> label_countries


```


```{r echo=F, message = F, warning = F, fig1, fig.height = 7, fig.width = 10}
  plot_data %>% ggplot(aes(x =Date,y = Cases,color=Country.Region)) + geom_path() + geom_point() + scale_x_date(date_labels = "%b/%d")  + theme_economist(base_size = 16)  +  guides(color = FALSE)   + geom_text_repel(data =plot_data %>% filter(Date == last(Date)), aes(label = label_countries), nudge_x =2,xlim = c(last(plot_data$Date),NA), hjust = 0, direction ='y',segment.size = 0.1, segment.colour = 'gray50',size = 4)  + xlim(as.Date('2020-02-14'), last(plot_data$Date) + 4 )

```

<br><br>

## Total cases per day since day of 100th case


```{r echo=F, message = F, warning = F, fig2, fig.height = 11, fig.width = 10 }



# find the date then substract to normalize the axis 
plot_data %>% group_by(Country.Region) %>% filter(Cases >99) %>% top_n(1,-Cases) -> plot_data_since_100

plot_data[,"Diff"] <- NA
for(i in 1:dim( plot_data_since_100 )[1] ){
   plot_data[plot_data$Country.Region==plot_data_since_100$Country.Region[i],]$Diff =  plot_data$Date[plot_data$Country.Region==plot_data_since_100$Country.Region[i]] - plot_data_since_100$Date[i]
    
}


plot_data %>% ggplot(aes(x =Diff,y = Cases,color=Country.Region)) + geom_path() + geom_point() + theme_economist(base_size = 16)  +  guides(color = FALSE)   + geom_text_repel(data =plot_data %>% filter(Date == last(Date)), aes(label = label_countries), nudge_x =2,xlim = c(max(plot_data$Diff),NA), hjust = 0, direction ='y',segment.size = 0.1, segment.colour = 'gray50',size = 4)  + xlim(0,max(plot_data$Diff)+4)


```


<br><br>

## Cases per million habitants (since day of 100th case)



```{r echo=F, message = F, warning = F, fig7, fig.width = 10}


world_moll %>% select(name,pop_est) -> world_population

world_population<-as.data.frame(world_population)

names(world_population)<-c("Country.Region","population","geometry")

world_population$population<-world_population$population/10^6
plot_data_pm = plot_data
plot_data_pm$Country.Region %>% str_replace("US","United States") %>% str_replace("Korea, South","Korea") -> plot_data_pm$Country.Region

left_join(plot_data_pm,world_population,by='Country.Region') %>% mutate(Cases_pm = Cases/population) ->plot_data_pm

 plot_data_pm$Diff = plot_data$Diff

 plot_data_pm %>% ggplot(aes(x =Diff,y = Cases_pm,color=Country.Region)) + geom_path() + geom_point() + theme_economist(base_size = 16)  +  guides(color = FALSE)   + geom_text_repel(data =plot_data_pm %>% filter(Date == last(Date)), aes(label = label_countries), nudge_x =2,xlim = c(max(plot_data_pm$Diff),NA), hjust = 0, direction ='y',segment.size = 0.1, segment.colour = 'gray50',size = 4)  + xlim(0,max(plot_data$Diff)+4)

```






<br><br>

## Total deaths per day


```{r echo=F, message = F, warning = F}
tidy_deaths %>% group_by(Country.Region,Date) %>% summarize(Cases = sum(Cases)) %>% filter(Cases ==max(Cases)) %>% arrange(desc(Cases)) %>% filter(Cases>=20) %>% select(Country.Region) -> plot_countries_deaths


tidy_deaths$Date <-as.Date(tidy_deaths$Date)

tidy_deaths %>% group_by(Country.Region,Date) %>% summarize(Cases = sum(Cases)) %>% filter(Country.Region %in% plot_countries_deaths$Country.Region)-> plot_deaths 

plot_deaths %>% filter(Date==last(Date)) %>% pull(Country.Region) %>% as.character() -> label_countries_deaths 

label_countries_deaths %>% str_replace(pattern = "United Kingdom","UK") %>% str_replace("Korea, South","S Korea") -> label_countries_deaths



```


```{r echo=F, message = F, warning = F, fig4, fig.height = 7, fig.width = 10}
plot_deaths %>% ggplot(aes(x =Date,y = Cases,color=Country.Region)) + geom_path() + geom_point() + scale_x_date(date_labels = "%b/%d")  + theme_wsj(base_size = 16)  +  guides(color = FALSE)   + geom_text_repel(data =plot_deaths %>% filter(Date == last(Date)), aes(label = label_countries_deaths), nudge_x =2,xlim = c(last(plot_deaths$Date),NA), hjust = 0, direction ='y',segment.size = 0.2, segment.colour = 'gray50',size = 4)  + xlim(first(plot_deaths$Date) + 4, last(plot_deaths$Date) + 6 )
```


<br><br>

## Total deaths per day since 10th death


```{r echo=F, message = F, warning = F, fig5, fig.height = 7, fig.width = 10 }

# find the date then substract to normalize the axis 
plot_deaths %>% group_by(Country.Region) %>% filter(Cases >10) %>% top_n(1,-Cases) -> plot_deaths_since_10

plot_deaths[,"Diff"] <- NA
for(i in 1:dim( plot_deaths_since_10 )[1] ){
   plot_deaths[plot_deaths$Country.Region==plot_deaths_since_10$Country.Region[i],]$Diff =  plot_deaths$Date[plot_deaths$Country.Region==plot_deaths_since_10$Country.Region[i]] - plot_deaths_since_10$Date[i]
    
}


plot_deaths %>% filter(Cases>10) %>% 
  ggplot(aes(x =Diff,y = Cases,color=Country.Region)) + geom_path() + geom_point() + theme_wsj(base_size = 16)  +  guides(color = FALSE)   + 
    geom_text_repel(data =plot_deaths %>% filter(Date == last(Date)), 
                    aes(label = label_countries_deaths), nudge_x =2,
                    xlim = c(max(plot_deaths$Diff[plot_deaths$Country.Region=='Iran']) + 7,NA), hjust = 0, direction ='y',segment.size = 0.2, segment.colour = 'gray50',size = 4)  + xlim(0,max(plot_deaths$Diff)+4) + ylab("Number of deaths (log)") + xlab("Days since 10th death") + theme(axis.title=element_text(size=20))    + scale_y_continuous(trans="log10")+ annotation_logticks(sides = 'l')


```

