---
title: "Basics 2: Exploring a data frame"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Titanic data set

When doing data analysis we need to get to know our data. What are the variables? What type of values do they have? And most importantly: What kind of questions can we ask?

In this section we will learn how to explore a dataset. We will use the `Titanic` dataset that can be easily installed in R. 

While exploring the dataset, we will also learn some useful function to explore, organize and visualize data. 

```{r cars}
#for fancy plots
library(ggplot2)
#install.packages("titanic")
library(titanic)

#For data frame manipulation
library(dplyr)

titanic<-titanic_train
head(titanic)
```

## Exploring the data

Explore the values for some variables:

```{r table}
View(titanic)

```

### Histogram

One of the most informative ways to explore a data set is by looking at the distribution of values for a given variable. A distribution tells us how the _observed_ values of the variable are spread across the range of _possible_ values.

```{r hist1, echo=T}
hist(titanic$Age,main="Age distribution for Titanic passengers",ylab ="Age (years)",xlab ="Number of passengers",col = "darkcyan")
```

We can use the command `table()` to explore get frequencies of each value in the data. Let's try for gender:
```{r sex, echo=T}
table(titanic$Sex)
table(titanic$Pclass)
table(titanic$Survived)
```

Some variables such as _Survived_ and _Sex_ have only two values, that can be numbers or words. In _R_ we should specify them as `factors` which is a type of variable that _R_ can understand and compare even words and numbers. 

We are going to use a for loop to convert 4 variables into factors. 

First, we create the array `cols` with the names of the variables. The _foor_ loop will iterate through them and use the `as.factor()` function to convert the column in the dataframe. Note how we use `titanic[,i]` to select that particular column. 

```{r convert, echo=T}

cols<-c("Survived","Pclass","Sex","Embarked")
for (i in cols){
  titanic[,i] <- as.factor(titanic[,i])
}

```

### Barplot


Great, now we are ready to explore some relations between variables. Let's start with Gender and Survival. Do women survided more than man?

The `ggplot` function from the `ggplot2` library allows us to create complex plots, this is just a simple example. For more information check out the documentation. 

```{r stackedBar1, echo=T}

ggplot(data=titanic,aes(x=Sex,fill=Survived))+geom_bar()

```

So, it seems that most men died ... and most women survived. But is not fair to compare directly because there were _more men_ in the Titanic. We can compare _relative_ numbers using the `position="fill"` option. 

```{r stackedBar2, echo=T}

ggplot(data=titanic,aes(x=Sex,fill=Survived)) +
    geom_bar(position="fill") + 
      ylab("Frequency")

```

Now things are more evident. Right? Let's check out survival rate vs ticket class. Do rich people survive more than poor people?

```{r stackedBar3, echo=T}
ggplot(data = titanic,aes(x=Pclass,fill=Survived))+
    geom_bar(position="fill")+
      ylab("Frequency")

```

As expected (?), passengers from the first class had the highest chance to survive. Try to play around with other variables, by changing the `aes` arguments `x` and `fill`

### Boxplot

Now, we want to know if there is a difference in age between men and women. We could do it in a similar way as above but the variable `Age` takes many different values, from babies to old men. We need a different type of plot. A boxplot, displays the distribution of values in a very convenient way. We can group it by gender using `x=Sex` and we want to plot the variable Age so we set `y=Age` the `fill=Sex` is to color it by gender. 

```{r stackedBar4, echo=T}

ggplot(data=titanic,aes(x=Sex,y=Age,fill=Sex)) + geom_boxplot()

```

We can check 3 variables at the same time: Class, Sex and Age

```{r stackedBar5, echo=T}

ggplot(data=titanic,aes(x=Pclass,y=Age,fill=Sex)) + geom_boxplot()

```

We see that older people, both male and female, occupied the first class while younger people took the third class. Also, we see a general trend that men were older than women _independently_ of the class. 

Exercise: Can you explain what this command is doing? How do you interpret the plots?

```{r multHist, echo=T}
ggplot(titanic,aes(x=Age,fill=Survived)) +
  theme_bw() + facet_wrap(Sex~Pclass)+geom_histogram(binwidth=5) + 
  labs(y="Survived",x="Age")

```


### Organising data

Finally, we can also explore the data by sorting and filtering. The `dplyr` package allows us to do this and other useful operations to the data. This is just a quick introduction, for more advance usage see https://datascienceplus.com/getting-started-with-dplyr-in-r-using-titanic-dataset/

The operator `%>%` is very useful to create _pipelines_ of commands.

Let's `filter` the passengers that were men then `arrange` them by age, from oldest to yougest, with `desc`. Finally, we `select` _only_ the age and name of the passengers. The last command `head(10)` is just for showing the first few data rows instead of the whole data.frame. With the argument =10, we print only 10 passengers.

```{r dplyr1, echo=T}
#look at it
titanic %>% filter(Sex=="male") %>% arrange(desc(Age)) %>% select(Age,Name) %>% head(10)

#looks good? Save it in a new data.frame

age_name_sorted <- titanic %>% filter(Sex=="male") %>% 
                    arrange(desc(Age)) %>% select(Age,Name) 
```


Now, let's group the data with `group_by` and use the `summarise` function to calculate the average age of men and women.

```{r dplyr2, echo=T}
#we need to set na.rm = T to  remove the NA values!
titanic %>% group_by(Sex) %>% summarise(average_age = mean(Age,na.rm=T))

```

Now let's double group by Sex and Survive

```{r dplyr3, echo=T}
#we need to set na.rm = T to  remove the NA values!
titanic %>% group_by(Sex,Survived) %>% summarise(average_age = mean(Age,na.rm=T))

```

Finally, can you understand what this pipeline is doing?

```{r dplyr4, echo=T}

titanic %>%  mutate(Age_Bracket = ifelse(Age < 18, 'Minor','Major')) %>%
  group_by(Survived,Age_Bracket) %>%  
  summarise(pnt = n())

```

For more information and examples of how to use ggplot2 and dplyr please see the following references: 

## Dplyr

https://datascienceplus.com/getting-started-with-dplyr-in-r-using-titanic-dataset/

## ggplot2

https://www.r-graph-gallery.com/265-grouped-boxplot-with-ggplot2/

https://www.kaggle.com/hillabehar/titanic-analysis-with-r

https://towardsdatascience.com/data-analysis-and-visualisations-using-r-955a7e90f7dd